---
title: "Clustering and classification"
author: "Suvi"
date: "November 23, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Clustering and classification
<br>

```{r results='hide', message=FALSE, warning=FALSE}
setwd("~/yhis-kouluhommat/IODS-project")

library(ggplot2)
library(dplyr)
library(GGally)
library(MASS)
library(tidyr)
library(boot)
library(corrplot)

```


```{r}
data(Boston)
str(Boston)
summary(Boston)
```


<br>
In the Boston dataset, there are 506 observations and 14 variables, which have been studied in relation to housing values in suburbs of Boston, e.g. crime rate, taxation, population composition, and accessibility to radial highways.
<br><br><br>

<b> Variable frequencies and correlations </b>
<br>
```{r message=FALSE, warning=FALSE}

gather(Boston) %>% ggplot(aes(value)) + facet_wrap("key", scales = "free") + geom_bar()

```

<br>Crime rate (crim) average is 3.6 per capita. However, the distribution is rather skewed, since the maximum value is 88. Mean age of the sample is approximately 68 years. The tax rate variable is also very skewed to the right, i.e. the mode value of property-tax rate per \$10,000 is 650, but there is a low variablitiy on the low end as well (200-400). Similar distribution is seen in the accessibility to radial highways (rad variable)
<br><br><br>

```{r}
# calculate the correlation matrix and round it
cor_matrix<-cor(Boston) %>% round(digits=2)

# print the correlation matrix
cor_matrix

# visualize the correlation matrix
corrplot(cor_matrix, method="circle", type="upper", cl.pos = "b", tl.pos = "d",tl.cex = 0.6)

```


<br><br> Correlation matrix shows several large negative and positive associations. For example, the correlation between crime rate and accessibility to radial highways and tax-rate are 0.63 and 0.58 respectively. There is a strong negative correlation between proportion of non-retail business acres per town and weighted mean of distances to five Boston employment centres, and lower status of the population and median value of homes.
<br><br><br>
```{r}
# center and standardize variables
boston_scaled <- scale(Boston)

# summaries of the scaled variables
summary(boston_scaled)

# class of the boston_scaled object
class(boston_scaled)

# change the object to data frame
boston_scaled <- as.data.frame(boston_scaled)



```


<br><br> Here we have standardized all each observation of every variable by exctracting the mean and multiplying by the SD of said variable. The outcome is that all these variables have a mean of 0 and SD of 1.
<br><br><br>
<br> <b>Modifying the target variable </b>

```{r}
# class of the boston_scaled object
class(boston_scaled)


# summary of the scaled crime rate
summary(boston_scaled$crim)

# create a quantile vector of crim and print it
bins <- quantile(boston_scaled$crim)
bins

labels <- c("low","med_low","med_high","high")

# create a categorical variable 'crime'
crime <- cut(boston_scaled$crim, breaks = bins, include.lowest = TRUE, label=labels)


# look at the table of the new factor crime
table(crime)

# remove original crim from the dataset
boston_scaled <- dplyr::select(boston_scaled, -crim)

# add the new categorical value to scaled data
boston_scaled <- data.frame(boston_scaled, crime)


```


<br><br>
<b> Create test and train set </b>

```{r}
# number of rows in the Boston dataset 
n <- nrow(boston_scaled)

# choose randomly 80% of the rows
ind <- sample(n,  size = n * 0.8)

# create train set
train <- boston_scaled[ind,]

# create test set 
test <- boston_scaled[-ind,]

# save the correct classes from test data
correct_classes <- test$crime

# remove the crime variable from test data
test <- dplyr::select(test, -crime)

```

<br><br>
<b> Linear discriminant analysis </b>

```{r}
# linear discriminant analysis
lda.fit <- lda(crime ~., data = train)

# print the lda.fit object
lda.fit

# the function for lda biplot arrows
lda.arrows <- function(x, myscale = 1, arrow_heads = 0.1, color = "red", tex = 0.75, choices = c(1,2)){
  heads <- coef(x)
  arrows(x0 = 0, y0 = 0, 
         x1 = myscale * heads[,choices[1]], 
         y1 = myscale * heads[,choices[2]], col=color, length = arrow_heads)
  text(myscale * heads[,choices], labels = row.names(heads), 
       cex = tex, col=color, pos=3)
}

# target classes as numeric
classes <- as.numeric(train$crime)

# plot the lda results
plot(lda.fit, dimen = 2, col=classes, pch=classes)
lda.arrows(lda.fit, myscale = 3)

```


<br> LDA on the training set showed 3 LD dimensions. However, only the first LD seems to work in discriminating cases adequately.

<br><br>
```{r}
# predict classes with test data
lda.pred <- predict(lda.fit, newdata = test)

# cross tabulate the results
table(correct = correct_classes, predicted = lda.pred$class)

```

<br><br> The model seemed to be rather good at predicting high crime rate cases. The low-and in-between cases were not classified as accurately.


<br>

```{r}

# center and standardize variables
boston_scale <- scale(Boston)

# euclidean distance matrix
dist_eu <- dist(boston_scale)

# look at the summary of the distances
summary(dist_eu)



```

<br> The output shows that the average distance between the rows of a data matrix is 4.9.
<br><br>

```{r}
# k-means clustering
km <-kmeans(boston_scale, centers = 3)

# plot the Boston dataset with clusters
pairs(boston_scale, col = km$cluster)

```
<br>

```{r}
set.seed(123)

# determine the number of clusters
k_max <- 10

# calculate the total within sum of squares
twcss <- sapply(1:k_max, function(k){kmeans(boston_scale, k)$tot.withinss})

# visualize the results
qplot(x = 1:k_max, y = twcss, geom = 'line')


```
<br><br>
The nuber of clusters can be determined based on large drops in total within sum of squares. My interpretation is that there could be 2 to 3 clusters, but let's stick with 2.

<br>
```{r}
# k-means clustering
km <-kmeans(boston_scale, centers = 2)

# plot the Boston dataset with clusters
pairs(boston_scale, col = km$cluster)

km$size

center <-km$centers
center

```



<br> The output shows two clusters, the first is smaller than the second. 

<br> Cluster 1 is associated wih lower crime, average age, taxes, and proportion of non-retail business acres. It also has higher proportion of black people and distances to five Boston employment centres.

<br> Cluster 2 is associates with higher crime, nitrogen oxides concentration, age, accessibility to radial highways, taxes, and with more individuals with lower status of the population (percent).

<br><br><br>
